Spfy: an integrated graph-based sequence database for real-time prediction of Escherichia coli phenotypes

### 2018-01-30 paper-webserver.tex
ADD: new figures
CHANGE: shorten the intro (& move some of its material to other sections

\title{Spfy: applying a graph-based sequence database for real-time prediction of Escherichia coli phenotypes}

@@

% 3. 100,000 genomes, (lookup number for Enterobase, GenBank), how do we run analysis on it
WGS data for bacterial pathogens of public health importance have recently accumulated in public databases in the tens of thousands, with hundreds of thousands set to be available within the next few years. For \textit{Escherichia coli}, there are over sixty thousand publicly available genomes in EnteroBase \url{https://enterobase.warwick.ac.uk/} and three million whole genomes in GenBank \cite{doi:10.1093/nar/gks1195}.

To begin to solve this "big-data" problem, platforms such as BAP and IRIDA provide hosted solutions that compute results in real-time, and distribute analyses across computing resources. While effective for self-contained workflows, many of the comparative analyses that are run are broadly useful, and therefore computed multiple times. An effective method to mitigate the recomputing of analyses is to make the storage and retrieval of results part of the platform, and effectively linked to the organisms of interest with a standardized ontology. Such measures can help ensure the rapid response times required for public health applications, and allow results to be integrated and progressively updated as new data becomes available.

@@

% 5. how we solved it
In this study, we present the Spfy platform, which was built with a graph database at its core. Spfy aims to solve the problem of real-time analysis and merging of results from different analysis methods. By merging results and resolving connections between genome data, Spfy is able to identify relationships between all genomes sequenced in the past, present, and future. This graph-based result storage allows retrospective comparisons as more genomes are sequenced or populations change, and is flexible to accommodate new analysis methods as they are developed.

@@

By supporting multiple \textit{in-silico} subtyping options, the platform functions similar to a reference laboratory, while adding support for big-data analyses. Subtyping options for E.coli are O-antigen, H-antigen, Shiga-toxin 1 (Stx1), Shiga-toxin 2 (Stx2), and Intimin typing, VF and AMR annotation, pan-genome generation, group comparisons via Fisher's, and machine-learning (ML) based predictions for OmniLog AMR assays. These tasks are divided into subtasks, and distributed across a built-in task queue. Results are also decomposed into graph structures and stored within a larger graph database. By integrating task distribution with graph storage, Spfy enables large-scale analyses, such as epidemiological associations between specific genotypes, biomarkers, host, source, and other metadata, and statistical significance testing of genome markers for user-defined groups, in real-time. Currently, the platform has been tested with XXX genome files and result storage for XX analyses modules.

Future work will focus on adding additional analyses modules, using machine learning and artificial neural networks to aid genotype to phenotype predictions, and supporting different species. While the integrated approach of storing and retrieving results provides enormous benefits, the developed analyses modules are self-contained and can easily be integrated into existing platforms such as Galaxy \cite{goecks2010galaxy}, and IRIDA. The website is available at \url{https://lfz.corefacility.ca/superphy/spfy/}. Spfy's codebase is provided at \url{https://github.com/superphy/backend} and a developer guide is provided at \url{https://superphy.readthedocs.io/en/latest/}.

@@

@@

>>>>

\title{Spfy: an integrated graph-based sequence database for real-time prediction of Escherichia coli phenotypes}

@@

% 3. 100,000 genomes, (lookup number for Enterobase, GenBank), how do we run analysis on it
Traditionally, these platforms have been applied to samples numbering only a few dozens while WGS data for bacterial pathogens of public health importance have recently accumulated in public databases in the tens of thousands, with hundreds of thousands set to be available within the next few years. For \textit{Escherichia coli}, there are over sixty thousand publicly available genomes in EnteroBase \url{https://enterobase.warwick.ac.uk/} and three million whole genomes in GenBank \cite{doi:10.1093/nar/gks1195}.

Many of the comparative analyses that are run are broadly useful, and therefore computed multiple times. To begin to compute results on all available genomes, an effective method to mitigate the recomputing of analyses is to make the storage and retrieval of results part of the platform, and effectively linked to the organisms of interest with a standardized ontology. Such measures can help ensure the rapid response times required for public health applications, and allow results to be integrated and progressively updated as new data becomes available.

@@

% 5. how we solved it
In this study, we present the Spfy platform, which integrates a graph database with real-time analysis options. By integrating analysis options with data storage, we avoid recomputing analysis for identical runs. Graph-based result storage also allows retrospective comparisons as more genomes are sequenced or populations change, and is flexible to accommodate new analysis modules as they are developed. The website is available at \url{https://lfz.corefacility.ca/superphy/spfy/}.

@@

@@

By supporting multiple \textit{in-silico} subtyping options, the platform functions similar to a reference laboratory, while adding support for population analyses. Subtyping options for E.coli are O-antigen, H-antigen, Shiga-toxin 1 (Stx1), Shiga-toxin 2 (Stx2), and Intimin typing, VF and AMR annotation, pan-genome generation, group comparisons via Fisher's, and machine-learning (ML) based predictions for OmniLog AMR assays. These tasks are divided into subtasks, and distributed across a built-in task queue. Results are also decomposed into graph structures and stored within a larger graph database. By integrating task distribution with graph storage, Spfy enables large-scale analyses, such as epidemiological associations between specific genotypes, biomarkers, host, source, and other metadata, and statistical significance testing of genome markers for user-defined groups, in real-time. Currently, the platform has been tested with XXX genome files and result storage for XX analyses modules.

@@

Future work will focus on adding additional analyses modules, using machine learning and artificial neural networks to aid genotype to phenotype predictions, and supporting different species. While the integrated approach of storing and retrieving results provides enormous benefits, the developed analyses modules are self-contained and can easily be integrated into existing platforms such as Galaxy \cite{goecks2010galaxy}, and IRIDA. Spfy's codebase is provided at \url{https://github.com/superphy/backend} and a developer guide is provided at \url{https://superphy.readthedocs.io/en/latest/}.
